---
title: "LPS MIA accute OH2.6"
author: "Raphael Kübler"
date: "04/22/2021"
output:
  rmarkdown::html_document:
   theme: united
#   runtime: shiny
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r start, echo=TRUE, message=FALSE, results='hide'}
knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

LPS-MIA base analysis
```{r load_packages, echo=TRUE, message=FALSE, results='hide'}
library(DESeq2)
library(xlsx)
library(Hmisc)
library(dplyr)
library(viridis)
library(pheatmap)
library(variancePartition)
library(RColorBrewer)
library(pcaExplorer)
library(vsn)
library(EnhancedVolcano)
library(rapport)
library(factoextra)
library(broom)

options(check.names = FALSE)
```

## Load data
```{r load_data, echo=TRUE}
dir <- file.path("D:", "R", "R workspace")
setwd(dir)

load("04222021_LPS orgs_base.RData")

```

## Update functions
```{r functions, echo=TRUE}
PCA.covar <- function(df, meta, heatmap.r=F, data.r=F, heatmap.p=T,type="spearman"){
  l <- length(colnames(meta))
  l2 <- l+1
  pca.cov <- prcomp(t(df))
  PC.covs <- pca.cov$x[,1:l]
  PC_cov_correlation <- rcorr(as.matrix(PC.covs), as.matrix(meta), type=type)
  PC_variance.explained <- PC_cov_correlation$r[1:l,l2:length(rownames(PC_cov_correlation$r))]
  PC_cov.cor.p <- PC_cov_correlation$P[1:l,l2:length(rownames(PC_cov_correlation$r))]
  if (heatmap.r)
    pheatmap(PC_variance.explained, cluster_rows = F, cluster_cols = F, show_colnames = T,
             color = heatmap.color.code)
  if (heatmap.p)
    pheatmap(PC_cov.cor.p, cluster_rows = F, cluster_cols = F, show_colnames = T,
             color = colorRampPalette(c("red", "white"))(50),
             display_numbers = T)
  if (data.r)
    return (PC.covs)
}

createDT <- function(DF, caption="", scrollY=500){
data <- DT::datatable(DF, caption=caption,
extensions = 'Buttons',
options = list( dom = 'Bfrtip',
buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,
columnDefs = list(list(className = 'dt-center', targets = "_all"))
)
)
return(data)
}

```


## First step QC

(0) Subsetting and QC

```{r pre-processing, echo=TRUE, fig.height = 4, fig.width = 6, fig.align = "center"}
#Subsetting a df to samples of interest (accute)
subset.df <- gfilt.df[colnames(gfilt.df) %in% rownames(filt.meta[filt.meta$`Cell line` %in% c("OH2.6") & filt.meta$`Accute/chronic` %in% c("Accute"),])]
subset.df <- subset.df[rowSums(subset.df) >= length(colnames(subset.df)),] #Also removes ERCC SIs

subset.meta <- filt.meta_clean[rownames(filt.meta_clean) %in% colnames(subset.df),]

#Sort columns
subset.meta <- subset.meta[sort(rownames(subset.meta)),]
subset.df <- subset.df[sort(colnames(subset.df))]

all(rownames(subset.meta) == colnames(subset.df))

#Quick density QC
#550 x 450
plot.density(subset.meta,to.color="Stimulation",to.plot="`Library size`")
```

```{r pre-processing.1, echo=TRUE, fig.height = 5, fig.width = 6, fig.align = "center"}
#Quick covariate analysis
cov_for.cor <- cov_for.cor[sort(rownames(cov_for.cor)),]
cov_for.cor.sub <- cov_for.cor[rownames(cov_for.cor) %in% colnames(subset.df),]
cov_for.cor.sub <- cov_for.cor.sub[!colnames(cov_for.cor.sub) %in% c("Accute/chronic", "Cell line", "Batch", "Experiment")]
covar.cor.sub <- rcorr(as.matrix(cov_for.cor.sub), type="pearson")
covar.cor.sub_1 <- covar.cor.sub$r

#500x450
pheatmap(covar.cor.sub_1, color = heatmap.color.code)
```


```{r pre-processing.2, echo=TRUE, fig.height = 4, fig.width = 5, fig.align = "center"}

#D100 PCA-covariate analysis
#500x500
D100.PCAcovar <- PCA.covar(subset.df,cov_for.cor.sub,data.r=T,type="pearson")
PCA_cov_cor_R(cov_for.cor.sub, subset.df)
```


## Normalization

(1) Performing normalization

```{r normalization, echo=TRUE}
#For correcting the data
subset.meta$'Log(library size)' <- scale(subset.meta$`Library size`)
subset.meta$`Scaled percent. MT` <- scale(subset.meta$"Percent. MT")

#Create a DESeq2 matrix
dds.complex <- DESeqDataSetFromMatrix(countData = as.matrix(subset.df),
                                      colData = data.frame(subset.meta),
                                      design = ~ Log.library.size. + Percent..ERCC + Stimulation)


#version 2 based on dds object
# Estimate library size correction scaling factors
dds.complex <- estimateSizeFactors(dds.complex)

vsd.complex <- vst(dds.complex)

```

## Post-normalization QC

(2) Performing post-normalization QC:

```{r norm.qc, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
#Doing QC on the vst-transformed values
meanSdPlot(assay(vsd.complex))

sampleDists <- dist(t(assay(vsd.complex)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(vsd.complex)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

annot_df <- data.frame(colData(dds.complex)$Stimulation)
colnames(annot_df) <- c("Stimulation")
rownames(annot_df) <- colnames(dds.complex)

annot_colors <- vector('list', length=1)

annot_colors$Stimulation[["CTR"]] <- "#00AFBB"
annot_colors$Stimulation[["LPS"]] <- "#E7B800"
```

```{r norm.qc1, echo=TRUE, fig.height = 4, fig.width = 6.5, fig.align = "center"}
#Heatmap intersample-distance
pheatmap(sampleDistMatrix, 
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors,annotation_row = annot_df, annotation_colors = annot_colors)

```

```{r norm.qc2, echo=TRUE, fig.height = 5, fig.width = 4, fig.align = "center"}
#IQR of samples
norm.df <- assay(vsd.complex)
sOutliers.norm <- IQRoutlier.detector(norm.df,barplot=T)

```

```{r norm.qc3, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
#PCA
#550x450
pca.norm <- plotPCA.custom(as.matrix(norm.df), intgroup=c("Cell line", "Stimulation","Experiment"), 
                           ntop = 50000, returnData=TRUE, metadata=subset.meta)
PoV.norm <- round(100 * attr(pca.norm, "percentVar"))
PCAplot(pca.norm, Condition="Stimulation", PoV.df=PoV.norm)


```

## Data correction

(3) Correcting the data

```{r plots.1, echo=TRUE}
#Remove batch effects
#Create a meta df that corresponds to the sorted columns of the vsd assay but has covariates as numeric factors
covs_forremBatch <- subset.meta
covs_forremBatch$Stimulation <- as.factor(as.numeric(covs_forremBatch$Stimulation))
covs_forremBatch$Experiment <- as.factor(as.numeric(covs_forremBatch$Experiment))
covs_forremBatch$`Cell line` <- as.factor(as.numeric(covs_forremBatch$`Cell line`))
covs_forremBatch$`Batch` <- as.factor(as.numeric(covs_forremBatch$`Batch`))
covs_forremBatch$Replicate <- as.factor(as.numeric(as.factor(covs_forremBatch$Replicate)))
covs_forremBatch <- covs_forremBatch[!colnames(covs_forremBatch) %in% c("Duplicate","Accute/chronic")]

batch.rem <- removeBatchEffect(as.matrix(assay(vsd.complex)), 
                               covariates=as.matrix(cbind(covs_forremBatch$`Log(library size)`,
                                                          covs_forremBatch$`Percent. ERCC`
                                                          )),
                               design=model.matrix(~ covs_forremBatch$Stimulation))


```

## Post-correction QC

(4) Post-correction QC

```{r QC.post, echo=TRUE, fig.height = 4, fig.width = 4, fig.align = "center"}

#====PC loadings for covariates====#
#500x500
covs_forremBatch <- covs_forremBatch[colnames(covs_forremBatch) %in% colnames(cov_for.cor.sub)]
D100.PCAcovar.postcorrection <- PCA.covar(batch.rem,covs_forremBatch,data.r=T,type="pearson",heatmap.r=T)

#PCA (550 x 450)
pca.cor <- plotPCA.custom(as.matrix(batch.rem), intgroup=c("Stimulation","Cell line","Experiment"),
                          ntop = 50000, returnData=TRUE, metadata=subset.meta, pc.1=1,pc.2=2)
PoV.cor <- round(100 * attr(pca.cor, "percentVar"))
```

```{r norm.qc3.1, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
#600x450
PCAplot(pca.cor, "Stimulation",PoV.df=PoV.cor)

```


```{r QC.post2, echo=TRUE, fig.height = 5, fig.width = 5.5, fig.align = "center"}
#Inter sample distances post correction
sampleDists.postcor <- dist(t(batch.rem))
sampleDistMatrix.postcor <- as.matrix(sampleDists.postcor)
rownames(sampleDistMatrix.postcor) <- colnames(vsd.complex)
colnames(sampleDistMatrix.postcor) <- NULL


#Heatmap intersample-distance
#650x500
pheatmap(sampleDistMatrix.postcor, show_rownames = F,
         clustering_distance_rows=sampleDists.postcor,
         clustering_distance_cols=sampleDists.postcor,
         col=colors, annotation_row = annot_df, annotation_colors = annot_colors)

```


## DESeq2 anaylsis

(5) Performing DESeq2

```{r DESeq1, echo=TRUE}
dds.complex <- DESeq(dds.complex)

res.complex <- results(dds.complex, name="Stimulation_LPS_vs_CTR")

summary(res.complex)
```

## Interactive results table

(5.1) Results

```{r DESeq2, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#DT::renderDT(data.frame(res.complex), "OH1.5A",scrollY=1000)

createDT(data.frame(res.complex), "OH1.5A", scrollY=1000)

```

## Plotting results

(6) Plotting results

```{r plots.5, echo=TRUE, fig.height = 6, fig.width = 10, fig.align = "center"}
#Volcano plot
#700x750
EnhancedVolcano(res.complex,
                lab = rownames(res.complex),
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.05,
                FCcutoff = 2,
                labSize = 6,
                
                legendPosition = 'right')

```

```{r plots.6, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
#Grab significant downreg & highest lfc genes
if (length(res.complex[is.na(res.complex$padj),]$padj) != 0)
  res.complex[is.na(res.complex$padj),]$padj <- 1

downreg.genes <- rownames(res.complex[res.complex$log2FoldChange < -2 & res.complex$padj < 0.05,])
upreg.genes <- rownames(res.complex[res.complex$log2FoldChange > 2 & res.complex$padj < 0.05,])


#Plot the genes
#heatmap based on sorted samples for Stimulation:
fact.to.sort <- as.numeric(annot_df$Stimulation)
sorted <- batch.rem[,order(fact.to.sort, decreasing=F)]
sub <- sorted[rownames(sorted) %in% c(upreg.genes,downreg.genes),]
#550x1300
pheatmap(sub, 
         cluster_rows = T, 
         cluster_cols = F, annotation_legend = T, show_colnames = F,
         annotation_col=annot_df[-2], 
         color = heatmap.color.code, scale='row', annotation_colors = annot_colors)


```

## LPS response genes

(6.1) LPS response and microglia genes
```{r plots.6.1, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
LPS.genes <- sort(c("IFT1", "IFT2", "IFT3", "CXCL8", "CX3CR1", "IL10", "IL1B", "TNF", 
               "IL6", "TMEM119", "AIF1", "TLR4", "IL12A", "IL23A", "CD40", "P2RY12", "GFAP"))
LPS.sub <- sorted[rownames(sorted) %in% c(LPS.genes),]

pheatmap(LPS.sub, 
         cluster_rows = T, 
         cluster_cols = F, annotation_legend = T, show_colnames = F,
         annotation_col=annot_df[-2], 
         color = heatmap.color.code, scale='row', annotation_colors = annot_colors)


```

## Boxplots of DEGs

(7) Boxplots of DEGs

```{r plots.7, echo=TRUE, fig.height = 10, fig.width = 12, fig.align = "center"}
#Boxplots of DEGs
rownames(sub) <- gsub('\\.', '_', rownames(sub))
rownames(sub) <- gsub('\\-', '_', rownames(sub))
box <- data.frame(scale(t(sub)), Stimulation=annot_df[order(fact.to.sort, decreasing=F),])


#box.chr <- data.frame(t(sub.ac), Stimulation=sub.meta.ac$Stimulation)
box$Stimulation <- factor(box$Stimulation, levels=c("CTR","LPS"))
#Boxplots
library(ggplot2)

all.p <- NULL

for (i in rownames(sub)){
  n <- as.factor(i)
  p = 
    ggplot(box, aes_string(x="Stimulation", y=i, fill="Stimulation")) +
    geom_boxplot() +
    scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))+
    theme(
      legend.position="Type",
      plot.title = element_text(size=11)
    ) +
    theme_bw()+
    theme(plot.title = element_text(hjust=0.5))+
    ylab(paste(i,"(residual expression)"))+
    theme(axis.text.x=element_blank(), axis.title.x=element_blank())
  all.p[[i]] <- p}


library(ggpubr)
#700x550
all.genes <- ggarrange(plotlist=all.p[1:length(all.p)], common.legend = T)
all.genes

```


