---
title: "LPS MIA chronic all samples"
author: "Raphael Kübler"
date: "04/28/2021"
output:
  rmarkdown::html_document:
   theme: united
#   runtime: shiny
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r start, echo=TRUE, message=FALSE, results='hide'}
knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = TRUE,
  fig.show = "hold")
```

LPS-MIA base analysis
```{r load_packages, echo=TRUE, message=FALSE, results='hide'}
library(DESeq2)
library(xlsx)
library(Hmisc)
library(dplyr)
library(viridis)
library(pheatmap)
library(variancePartition)
library(RColorBrewer)
library(pcaExplorer)
library(vsn)
library(EnhancedVolcano)
library(rapport)
library(factoextra)
library(broom)
library(readr)
options(check.names = FALSE)
```

## Load data
```{r load_data, echo=TRUE}
dir <- file.path("D:", "R", "R workspace")
setwd(dir)

load("04222021_LPS orgs_base.RData")

```

## Update functions
```{r functions, echo=TRUE}
PCA.covar <- function(df, meta, heatmap.r=F, data.r=F, heatmap.p=T,type="spearman"){
  l <- length(colnames(meta))
  l2 <- l+1
  pca.cov <- prcomp(t(df))
  PC.covs <- pca.cov$x[,1:l]
  PC_cov_correlation <- rcorr(as.matrix(PC.covs), as.matrix(meta), type=type)
  PC_variance.explained <- PC_cov_correlation$r[1:l,l2:length(rownames(PC_cov_correlation$r))]
  PC_cov.cor.p <- PC_cov_correlation$P[1:l,l2:length(rownames(PC_cov_correlation$r))]
  if (heatmap.r)
    pheatmap(PC_variance.explained, cluster_rows = F, cluster_cols = F, show_colnames = T,
             color = heatmap.color.code)
  if (heatmap.p)
    pheatmap(PC_cov.cor.p, cluster_rows = F, cluster_cols = F, show_colnames = T,
             color = colorRampPalette(c("red", "white"))(50),
             display_numbers = T)
  if (data.r)
    return (PC.covs)
}

createDT <- function(DF, caption="", scrollY=500){
data <- DT::datatable(DF, caption=caption,
extensions = 'Buttons',
options = list( dom = 'Bfrtip',
buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,
columnDefs = list(list(className = 'dt-center', targets = "_all"))
)
)
return(data)
}

```


## First step QC

(0) Subsetting and QC

```{r pre-processing, echo=TRUE, fig.height = 4, fig.width = 6, fig.align = "center"}
#Subsetting a df to samples of interest (accute)
subset.df <- gfilt.df[colnames(gfilt.df) %in% rownames(filt.meta[filt.meta$`Accute/chronic` %in% c("Chronic"),])]
subset.df <- subset.df[rowSums(subset.df) >= length(colnames(subset.df)),] #Also removes ERCC SIs

subset.meta <- filt.meta_clean[rownames(filt.meta_clean) %in% colnames(subset.df),]

#Sort columns
subset.meta <- subset.meta[sort(rownames(subset.meta)),]
subset.df <- subset.df[sort(colnames(subset.df))]

all(rownames(subset.meta) == colnames(subset.df))

#Quick density QC
#550 x 450
plot.density(subset.meta,to.color="Stimulation",to.plot="`Library size`")

#Quick covariate analysis
cov_for.cor <- cov_for.cor[sort(rownames(cov_for.cor)),]
cov_for.cor.sub <- cov_for.cor[rownames(cov_for.cor) %in% colnames(subset.df),]
cov_for.cor.sub <- cov_for.cor.sub[!colnames(cov_for.cor.sub) %in% c("Accute/chronic")]
covar.cor.sub <- rcorr(as.matrix(cov_for.cor.sub), type="pearson")
covar.cor.sub_1 <- covar.cor.sub$r

#500x450
pheatmap(covar.cor.sub_1, color = heatmap.color.code)
```


```{r pre-processing.2, echo=TRUE, fig.height = 4, fig.width = 5, fig.align = "center"}

#D100 PCA-covariate analysis
#500x500
D100.PCAcovar <- PCA.covar(subset.df,cov_for.cor.sub,data.r=T,type="pearson")
#PCA_cov_cor_R(cov_for.cor.sub, subset.df)
```


## Normalization

(1) Performing normalization

```{r normalization, echo=TRUE}
#For correcting the data
subset.meta$'Log(library size)' <- scale(subset.meta$`Library size`)
subset.meta$`Scaled percent. MT` <- scale(subset.meta$"Percent. MT")

#Create a DESeq2 matrix
dds.complex <- DESeqDataSetFromMatrix(countData = as.matrix(subset.df),
                                      colData = data.frame(subset.meta),
                                      design = ~ Experiment + Log.library.size. + Percent..ERCC + Scaled.percent..MT + Stimulation)


#version 2 based on dds object
# Estimate library size correction scaling factors
dds.complex <- estimateSizeFactors(dds.complex)

vsd.complex <- vst(dds.complex)

```

## Post-normalization QC

(2) Performing post-normalization QC:

```{r norm.qc1, echo=TRUE, fig.height = 4, fig.width = 6.5, fig.align = "center"}
#Doing QC on the vst-transformed values
meanSdPlot(assay(vsd.complex))

sampleDists <- dist(t(assay(vsd.complex)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(vsd.complex)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

annot_df <- data.frame(colData(dds.complex)$Stimulation,colData(dds.complex)$`Cell.line`)
colnames(annot_df) <- c("Stimulation","Cell line")
rownames(annot_df) <- colnames(dds.complex)

annot_colors <- vector('list', length=1)

annot_colors$Stimulation[["CTR"]] <- "#00AFBB"
annot_colors$Stimulation[["LPS"]] <- "#E7B800"

annot_colors$`Cell line`[["OH1.5"]] <- "#2E7281"
annot_colors$`Cell line`[["OH2.6"]] <- "black"
annot_colors$`Cell line`[["PRO1"]] <- "#50B8CF"

#Heatmap intersample-distance
pheatmap(sampleDistMatrix, 
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors,annotation_row = annot_df, annotation_colors = annot_colors)

```

```{r norm.qc2, echo=TRUE, fig.height = 5, fig.width = 4, fig.align = "center"}
#IQR of samples
norm.df <- assay(vsd.complex)
sOutliers.norm <- IQRoutlier.detector(norm.df,barplot=T)

```

```{r norm.qc3, echo=TRUE, fig.height = 4, fig.width = 6.5, fig.align = "center"}
#PCA
#550x450
pca.norm <- plotPCA.custom(as.matrix(norm.df), intgroup=c("Cell line", "Stimulation","Experiment"), 
                           ntop = 50000, returnData=TRUE, metadata=subset.meta)
PoV.norm <- round(100 * attr(pca.norm, "percentVar"))
PCAplot(pca.norm, Condition="Stimulation", PoV.df=PoV.norm)
PCAplot(pca.norm, "Cell line", Shape="Experiment",PoV.df=PoV.norm)

```

## Data correction

(3) Correcting the data

```{r plots.1, echo=TRUE}
#Remove batch effects
#Create a meta df that corresponds to the sorted columns of the vsd assay but has covariates as numeric factors
covs_forremBatch <- subset.meta
covs_forremBatch$Stimulation <- as.factor(as.numeric(covs_forremBatch$Stimulation))
covs_forremBatch$Experiment <- as.factor(as.numeric(covs_forremBatch$Experiment))
covs_forremBatch$`Cell line` <- as.factor(as.numeric(covs_forremBatch$`Cell line`))
covs_forremBatch$`Batch` <- as.factor(as.numeric(covs_forremBatch$`Batch`))
covs_forremBatch$Replicate <- as.factor(as.numeric(as.factor(covs_forremBatch$Replicate)))
covs_forremBatch <- covs_forremBatch[!colnames(covs_forremBatch) %in% c("Duplicate","Accute/chronic")]

batch.rem <- removeBatchEffect(as.matrix(assay(vsd.complex)), batch = covs_forremBatch$Experiment, 
                               covariates=as.matrix(cbind(covs_forremBatch$`Log(library size)`,
                                                          covs_forremBatch$`Percent. ERCC`,covs_forremBatch$`Scaled percent. MT`
                                                          )),
                               design=model.matrix(~ covs_forremBatch$Stimulation))
batch.rem2 <- removeBatchEffect(as.matrix(assay(vsd.complex)), batch = covs_forremBatch$Experiment, batch2 = covs_forremBatch$Replicate,
                               covariates=as.matrix(cbind(covs_forremBatch$`Log(library size)`,
                                                          covs_forremBatch$`Percent. ERCC`,covs_forremBatch$`Scaled percent. MT`
                                                          )),
                               design=model.matrix(~ covs_forremBatch$Stimulation))


```

## Post-correction QC

(4) Post-correction QC

```{r QC.post, echo=TRUE, fig.height = 4, fig.width = 4, fig.align = "center"}

#====PC loadings for covariates====#
#500x500
covs_forremBatch <- covs_forremBatch[colnames(covs_forremBatch) %in% colnames(cov_for.cor.sub)]
D100.PCAcovar.postcorrection <- PCA.covar(batch.rem,covs_forremBatch,data.r=T,type="pearson",heatmap.r=T)
#PCA_cov_cor_R(covs_forremBatch, batch.rem)

#PCA (550 x 450)
pca.cor <- plotPCA.custom(as.matrix(batch.rem), intgroup=c("Stimulation","Cell line","Experiment"),
                          ntop = 50000, returnData=TRUE, metadata=subset.meta, pc.1=1,pc.2=2)
PoV.cor <- round(100 * attr(pca.cor, "percentVar"))

```


```{r norm.qc3.1, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
#600x450
PCAplot(pca.cor, "Stimulation",PoV.df=PoV.cor)
PCAplot(pca.cor, "Cell line", Shape="Experiment",PoV.df=PoV.cor)

```

Something that might make you sad: this is how the data looks when we can remove duplicate correlation:
```{r norm.qc3.2, echo=TRUE, fig.height = 4, fig.width = 6.5, fig.align = "center"}
pca.cor2 <- plotPCA.custom(as.matrix(batch.rem2), intgroup=c("Stimulation","Cell line","Experiment"),
                          ntop = 50000, returnData=TRUE, metadata=subset.meta, pc.1=1,pc.2=2)
PoV.cor2 <- round(100 * attr(pca.cor2, "percentVar"))
PCAplot(pca.cor2, "Stimulation",PoV.df=PoV.cor2)
PCAplot(pca.cor2, "Cell line", Shape="Experiment",PoV.df=PoV.cor2)

```


```{r QC.post2, echo=TRUE, fig.height = 5, fig.width = 5.2, fig.align = "center"}
#Inter sample distances post correction
sampleDists.postcor <- dist(t(batch.rem))
sampleDistMatrix.postcor <- as.matrix(sampleDists.postcor)
rownames(sampleDistMatrix.postcor) <- colnames(vsd.complex)
colnames(sampleDistMatrix.postcor) <- NULL


#Heatmap intersample-distance
#650x500
pheatmap(sampleDistMatrix.postcor, show_rownames = F,
         clustering_distance_rows=sampleDists.postcor,
         clustering_distance_cols=sampleDists.postcor,
         col=colors, annotation_row = annot_df, annotation_colors = annot_colors)

```


## DESeq2 anaylsis

(5) Performing DESeq2

```{r DESeq1, echo=TRUE}
dds.complex <- DESeq(dds.complex)

res.complex <- results(dds.complex, name="Stimulation_LPS_vs_CTR")

summary(res.complex)
```

## Interactive results table

(5.1) Results

```{r DESeq2, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#DT::renderDT(data.frame(res.complex), "OH1.5A",scrollY=1000)

createDT(data.frame(res.complex), "Accute", scrollY=1000)

```

## Plotting results

(6) Plotting results

```{r plots.5, echo=TRUE, fig.height = 6, fig.width = 10, fig.align = "center"}
#Volcano plot
#700x750
EnhancedVolcano(res.complex,
                lab = rownames(res.complex),
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.05,
                FCcutoff = 2,
                labSize = 6,
                
                legendPosition = 'right')

```

```{r plots.6, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
#Grab significant downreg & highest lfc genes
if (length(res.complex[is.na(res.complex$padj),]$padj) != 0)
  res.complex[is.na(res.complex$padj),]$padj <- 1

downreg.genes <- rownames(res.complex[res.complex$log2FoldChange < -2 & res.complex$padj,])
upreg.genes <- rownames(res.complex[res.complex$log2FoldChange > 2 & res.complex$padj,])


#Plot the genes
#heatmap based on sorted samples for Stimulation:
fact.to.sort <- as.numeric(annot_df$`Cell line`)
sorted <- batch.rem[,order(fact.to.sort, decreasing=F)]
sub <- sorted[rownames(sorted) %in% c(upreg.genes,downreg.genes),]
#550x1300
pheatmap(sub, 
         cluster_rows = T, 
         cluster_cols = F, annotation_legend = T, show_colnames = F,
         annotation_col=annot_df, 
         color = heatmap.color.code, scale='row', annotation_colors = annot_colors)


```

## LPS response & microglia genes

(6.1) LPS response genes
```{r plots.6.1, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
LPS.genes <- sort(c("IFT1", "IFT2", "IFT3", "CXCL8", "CX3CR1", "IL10", "IL1B", "TNF", 
               "IL6", "TMEM119", "AIF1", "TLR4", "IL12A", "IL23A", "CD40", "P2RY12", "GFAP"))
LPS.sub <- sorted[rownames(sorted) %in% c(LPS.genes),]

pheatmap(LPS.sub, 
         cluster_rows = T, 
         cluster_cols = F, annotation_legend = T, show_colnames = F,
         annotation_col=annot_df, 
         color = heatmap.color.code, scale='row', annotation_colors = annot_colors)


```


## Patir et al. microglia markers
(6.2) Patir et al. marker genes
```{r plots.6.2, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
dir <- file.path("C:", "Users", "rapha", "Dokumente", "UM FN", "Internship", "Internship_Start", "Project 4 - Synthesis", "Analysis", "Genelists")

setwd(dir)

coresign <- read_lines("PatirEtAl_coreMicrogliaGenes.csv")
genesOI <- batch.rem[rownames(batch.rem) %in% coresign,]

pheatmap(genesOI, 
         cluster_rows = F, show_rownames = F,
         cluster_cols = T, annotation_legend = T, show_colnames = F,
         annotation_col=annot_df, fontsize = 6,
         color = heatmap.color.code, scale='row', annotation_colors = annot_colors)

pheatmap(genesOI, 
         cluster_rows = F, show_rownames = F,
         cluster_cols = F, annotation_legend = T, show_colnames = F,
         annotation_col=annot_df,
         color = heatmap.color.code, scale='row', annotation_colors = annot_colors)



```


## Patir et al. microglia markers
(7) Composite microglia marker score:
```{r plots.7, echo=TRUE, fig.height = 4, fig.width = 4, fig.align = "center"}
Patir_score <- prcomp(t(genesOI))

#identical(colnames(genesOI), rownames(subset.meta))


#rcorr(as.matrix(Patir_score$x), as.matrix(as.numeric(subset.meta$Stimulation)))

PCA.covar(genesOI,covs_forremBatch,data.r=F,type="pearson",heatmap.r=T)

PCA_cov_cor_R(covs_forremBatch,genesOI)


```

(7.1) Look at the PCA:
```{r plots.71, echo=TRUE, fig.height = 4, fig.width = 6.5, fig.align = "center"}
pca.mic <- plotPCA.custom(as.matrix(genesOI), intgroup=c("Stimulation","Cell line","Experiment"),
                          ntop = 50000, returnData=TRUE, metadata=subset.meta, pc.1=1,pc.2=2)
PoV.mic <- round(100 * attr(pca.mic, "percentVar"))

PCAplot(pca.mic, "Stimulation", Shape="Experiment",PoV.df=PoV.mic)
```

## LPS response genes
(8) Composite LPS response score:
```{r plots.8, echo=TRUE, fig.height = 4, fig.width = 4, fig.align = "center"}
LPS_score <- prcomp(t(LPS.sub))

#identical(colnames(LPS.sub), rownames(subset.meta))

#rcorr(as.matrix(LPS_score$x), as.matrix(as.numeric(subset.meta$Stimulation)))

PCA.covar(LPS.sub,covs_forremBatch,data.r=F,T,type="pearson",heatmap.r=T)

#PCA_cov_cor_R(covs_forremBatch,LPS.sub)

```

```{r LPS.PCA, echo=TRUE, fig.height = 4, fig.width = 6.5, fig.align = "center"}

pca.LPS <- plotPCA.custom(as.matrix(LPS.sub), intgroup=c("Stimulation","Cell line","Experiment","Replicate"),
                          ntop = 50000, returnData=TRUE, metadata=subset.meta, pc.1=1,pc.2=2)
PoV.LPS <- round(100 * attr(pca.LPS, "percentVar"))

PCAplot(pca.LPS, "Stimulation", Shape="Experiment",PoV.df=PoV.LPS)

```


## Association between LPS response and microglia signatures
(9) How strongly are the microglia signature and the LPS response correlated?

```{r plots.9, echo=TRUE, fig.height = 5, fig.width = 5, fig.align = "center"}
LPS.mic <- data.frame(cbind("LPS"=scale(colMeans(LPS.sub)), "Microglia signature"=scale(colMeans(genesOI))))
colnames(LPS.mic) <- c("LPS","Microglia signature")
library(ggpmisc)
formula <- y ~ x  
ggplot(LPS.mic, aes_string(y= as.matrix(LPS.mic["LPS"]), x= as.matrix(LPS.mic["Microglia signature"]))) +  
  geom_point() + 
  scale_colour_gradient(low = "gray90", high = "black") +
  stat_smooth(method=lm) +
  theme_bw()+
  theme(plot.title = element_text(hjust=0.5))+
  geom_smooth(method=lm, linetype="dashed")+
  labs(y = "Cumulative LPS response score", x = "Microglia signature score")+
  stat_poly_eq(aes(label = c(paste(..adj.rr.label..))), 
               label.x = "right", label.y = 0.37,
               formula = formula, parse = TRUE, size = 3)+
  #  geom_cor(method='pearson')+
  stat_fit_glance(method = 'lm',
                  method.args = list(formula = formula),
                  geom = 'text',
                  aes(label = paste("P-value = ", signif(..p.value.., digits = 3), sep = "")),
                  label.x = 'right', label.y = 0.3, size = 3)

```
